---
title: "Red and White Wine Analysis"
output: html_document
---

# Red vs White wine quality exploration

## About the dataset

The datasets analysed about red and white wine were created by Paulo Cortez, Antonio Cerdeira, Fernando Almeida, Telmo Matos and Jose Reis, the wines studied are the Portuguese "Vinho Verde" in its red and white variant, downloaded from [here](https://docs.google.com/document/d/1qEcwltBMlRYZT-l699-71TzInWfk4W9q5rTCSvDVMpc/pub). The following variables were taken into consideration:

1. Fixed acidity (tartaric acid - g / dm^3) -- Most acids involved with wine or fixed or nonvolatile (do not evaporate readily).

2. Volatile acidity (acetic acid - g / dm^3) -- The amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste.

3. Citric acid (g / dm^3) -- Found in small quantities, citric acid can add 'freshness' and flavor to wines.

4. Residual sugar (g / dm^3) -- The amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet.

5. Chlorides (sodium chloride - g / dm^3) -- The amount of salt in the wine.

6. Free sulfur dioxide (mg / dm^3) -- Free sulfur dioxide: the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine.

7. Total sulfur dioxide (mg / dm^3) -- Amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine.

8. Density (g / cm^3) -- The density of water is close to that of water depending on the percent alcohol and sugar content.

9. pH -- Describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale.

10. Sulphates (potassium sulphate - g / dm3) -- A wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant.

11. Alcohol (% by volume) -- The percent alcohol content of the wine.

12. Quality (score between 0 and 10).

***

## Preparation of dataset

We start by loading the necessary libraries

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(GGally)
library(psych)
```

Loading the data and doing some data wrangling

```{r echo=FALSE, message=FALSE, warning=FALSE}
rw = read.csv("wineQualityReds.csv")
ww = read.csv("wineQualityWhites.csv")
rw = rw[,-1]
ww = ww[,-1]
rw$total.sulfur.dioxide.ppm = rw$total.sulfur.dioxide / 1000
ww$total.sulfur.dioxide.ppm = ww$total.sulfur.dioxide / 1000
rw$color = 'red'
ww$color = 'white'
wines = rbind(rw, ww)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
dim(rw)
dim(ww)
```
```{r}
summary(rw)
```

The dataset for red wine consists of 1599 observations and 13 variables, and the white wine dataset of 4898 observations and the same 13 variables. we will now analyse each variable separatedly.

***

## Single variable exploration

#### Fixed Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = fixed.acidity, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = fixed.acidity, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Fixed Acidity', y = 'Density')
summary(rw$fixed.acidity)
summary(ww$fixed.acidity)
```
The values of fixed acidity for red wine are larger than for white whine, most red wines have its acidity from 7 to 9 g/dm^3 of tartaric acid and most white wines between 6.3 and 6.3 g/dm^3 of tartaric acid. In both cases the distribution of the variables seem to present a normal behaviour.

***

### Volatile Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = volatile.acidity, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = volatile.acidity, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Volatile Acidity', y = 'Density')
summary(rw$volatile.acidity)
summary(ww$volatile.acidity)
```

Volatile acidity also follows a normal distribution for both types of wine, the range of values for white wines is smaller than the range for red wine, which contains most of them between 0.4 and 0.65. In addition to this it is evident that there is a greater concentration of smaller values of volatile acidity for white wine, while red whine has more variance.

***

#### Citric Acid

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = citric.acid, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = citric.acid, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Citric Acid', y = 'Density')
summary(rw$citric.acid)
summary(ww$citric.acid)
```
The data for citric acid concentrations for red wine does not appear normal, therefore a square root transformation is applied. Note that such transformation is possible due to the fact that there are no negative concentrations. Also, the mean value of citric acid for white wine, which does present normal behaviour is 0.33 g/dm^3.

***

#### Residual Sugar

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = residual.sugar, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = residual.sugar, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Residual Sugar', y = 'Density')
```
Given that the datasets have too many outliers, the right hand tail is insignificant, therefore we plot again the residual sugar subsetting the data to those in the inner 90 percent interval.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = subset(rw, residual.sugar < quantile(residual.sugar, 0.95) & residual.sugar >= quantile(residual.sugar, 0.05)), aes(x = residual.sugar, fill = "r"), alpha = 0.3) +
  geom_density(data = subset(ww, residual.sugar < quantile(residual.sugar, 0.95) & residual.sugar >= quantile(residual.sugar, 0.05)), aes(x = residual.sugar, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Residual Sugar', y = 'Density')
summary(subset(rw$residual.sugar, rw$residual.sugar < quantile(rw$residual.sugar, 0.95) & rw$residual.sugar >= quantile(rw$residual.sugar, 0.05)))
summary(subset(ww$residual.sugar, ww$residual.sugar < quantile(ww$residual.sugar, 0.95) & ww$residual.sugar >= quantile(ww$residual.sugar, 0.05)))
```
Here it is easy to note that white wine presents much more variance in Residual sugar, which is to be expected, while red wine have mostly between 1.9 and 2.5 g/dm^3 of residual sugar.

***

#### Chlorides

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = chlorides, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = chlorides, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Chlorides', y = 'Density')
```
Given that both datasets have too many outliers in the chlorides field, we plot again the residual sugar subsetting the data to those in the inner 90 percent interval.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = subset(rw, chlorides < quantile(chlorides, 0.95) & chlorides >= quantile(chlorides, 0.05)), aes(x = chlorides, fill = "r"), alpha = 0.3) +
  geom_density(data = subset(ww, chlorides < quantile(chlorides, 0.95) & chlorides >= quantile(chlorides, 0.05)), aes(x = chlorides, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Chlorides', y = 'Density')
summary(subset(rw$chlorides, rw$chlorides < quantile(rw$chlorides, 0.95) & rw$chlorides >= quantile(rw$chlorides, 0.05)))
summary(subset(ww$chlorides, ww$chlorides < quantile(ww$chlorides, 0.95) & ww$chlorides >= quantile(ww$chlorides, 0.05)))
```
Here we see that red wine has mean sodium chloride of 0.08 g/dm^3, twice as much than what is present in the white wine.

***

#### Free Sulfur Dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = free.sulfur.dioxide, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = free.sulfur.dioxide, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Free Sulfur Dioxide', y = 'Density')
summary(rw$free.sulfur.dioxide)
```
The range of values for free sulfur dioxide is very big, with a lenght of 71, however, the majority of values are in the range 7 to 21, this is due to the high density towards the cero, like a chi squared distribution. Therefore a sqrt transformation is applied.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = free.sulfur.dioxide, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = free.sulfur.dioxide, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Free Sulfur Dioxide (sqrt)', y = 'Density') +
  coord_trans(x = 'sqrt')
```

***

#### Total Sulfur Dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = total.sulfur.dioxide.ppm, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = total.sulfur.dioxide.ppm, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Total Sulfur Dioxide', y = 'Density')
summary(rw$total.sulfur.dioxide.ppm)
summary(ww$total.sulfur.dioxide.ppm)
```

As expected, total sulfur dioxide has a behavior similar to the free distribution measure, with a higher maximum value. A sqrt transformation is applied. It is of great importance to mantain the balance of total sulfur dioxide, due to the fact that when it surpluses concentrations of 50 ppm, SO2 becomes evident in the smell and taste of wine. For our samples this should not be a problem as the maximum value of SO2 is of 0.289ppm for the red wines and 0.44ppm for the white wines.

***

#### Density

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = density, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = density, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Density', y = 'Density')
summary(rw$density)
summary(ww$density)
```
To ignore outliers we set the plot to only represent values in the 90% range

```{r}
ggplot() + 
  geom_density(data = subset(rw, density < quantile(density, 0.95) & density >= quantile(density, 0.05)), aes(x = density, fill = "r"), alpha = 0.3) +
  geom_density(data = subset(ww, density < quantile(density, 0.95) & density >= quantile(density, 0.05)), aes(x = density, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Density', y = 'Density')
summary(rw$density)
summary(ww$density)
```
It is noticeable the difference in variance in the density between red and white wines, having 50% of their values between 0.9956 and 0.9978 g/cm^3 for red wines and  0.9917 and 0.9961 g/cm^3 for white wines.

***

#### pH

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = pH, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = pH, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'pH', y = 'Density')
summary(rw$pH)
summary(ww$pH)
```
No transformation will be applied to the variable pH. Red wines are more alcaline than white wines, having red wines mean pH of 3.311 and white wines 3.188.

***

#### Sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = sulphates, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = sulphates, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Sulphates', y = 'Density')
summary(rw$sulphates)
summary(ww$sulphates)
```
Sulphates have a behavior that seems exponential, therefore, a log10 transformation is applied to normalize the data.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = sulphates, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = sulphates, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Sulphates', y = 'Density') +
  coord_trans(x = 'log10')
```

***

#### Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = alcohol, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = alcohol, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Alcohol', y = 'Density')
summary(rw$alcohol)
summary(ww$alcohol)
```
The alcohol variable does not look much like a normal variable, however, it both varieties of wine seem to have the same percentages of alcohol by volume, varying between 8 and 15%.

***

#### Quality

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_density(data = rw, aes(x = quality, fill = "r"), alpha = 0.3) +
  geom_density(data = ww, aes(x = quality, fill = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Quality', y = 'Density')
summary(rw$quality)
summary(ww$quality)
```

Overall it seems white and red wines have the same densities of quality.

***

### Conclusion of the single variable exploration

Although most variables have normal behavior, a log10 transformation was necessary for sulphates and a sqrt transformation for free sulfur dioxide.

***

### What is the main feature of interest in the dataset?

The main features in the dataset are alcohol, residual sugar and quality. We would like to determine which features are best to predict the quality of a red wine. I suspect alcohol and residual sugar will be the most influential ones.

***

## Bivariate exploration

Now we study the correlation between the variables. To see if there is some pattern formed we will first look at a scatterplot matrix

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggpairs(wines[,c('quality', 'fixed.acidity', 'volatile.acidity','citric.acid', 'residual.sugar')], aes(color = wines$color, alpha = 0.5), lower = list(continuous = "smooth"))
ggpairs(wines[,c('quality', 'chlorides','free.sulfur.dioxide', 'total.sulfur.dioxide', 'density')], aes(color = wines$color, alpha = 0.5), lower = list(continuous = "smooth")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red"))
```

***

### Analysing alcohol vs quality

How much does alcohol percentage affect in quality? To study this we shall see the quality of those wines with alcohol within the middle 50% range and the quality for the outer 50% ranges.
  
**Middle 50% range:**
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(subset(rw$quality, rw$alcohol < quantile(rw$alcohol, 0.75) & rw$alcohol >= quantile(rw$alcohol, 0.25)))
summary(subset(rw$quality, rw$alcohol >= quantile(rw$alcohol, 0.75) | rw$alcohol < quantile(rw$alcohol, 0.25)))
```
Apparently alcohol does not affect the quality  much, however this might be due to the small range of alcohol percentages in our sample.


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot() + 
  geom_point(data = rw, aes(x = chlorides, y = alcohol, color = "r"), alpha = 0.3) +
  geom_point(data = ww, aes(x = chlorides, y = alcohol, color = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Chlorides', y = 'Alcohol')

ggplot() + 
  geom_point(data = rw, aes(x = pH, y = alcohol, color = "r"), alpha = 0.3) +
  geom_point(data = ww, aes(x = pH, y = alcohol, color = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'pH', y = 'Alcohol')

ggplot() + 
  geom_point(data = rw, aes(x = residual.sugar, y = alcohol, color = "r"), alpha = 0.3) +
  geom_point(data = ww, aes(x = residual.sugar, y = alcohol, color = "y"), alpha = 0.3) +
  scale_colour_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  scale_fill_manual(name ="Wine variety", values = c("r" = "#900c28", "y" = "#fdff8b"), labels=c("y" = "White", "r" = "Red")) +
  labs(x = 'Residual Sugar', y = 'Alcohol')

```

There seems to be a linearly decreasing relation between the amount of alcohol and the chlorides for both varieties of wine. **The residual sugar however does not seem to have enough variance to be worth studying**.




















